The Ky Fan Norms and Beyond: Dual Norms and Combinations for Matrix Optimization

### 2025-11-26 conf_article/neon.tex
added text about signsgd in sections 1 and 4

%------------------------------------------------------------------------------------
% \epsilon is predefined; override to use \varepsilon
\renewcommand{\epsilon}{\varepsilon}
\newcommand{\Rmn}{\R^{m\times n}}
\newcommand{\cB}{\mathcal{B}}
\newcommand{\cD}{\mathcal{D}}
\newcommand{\cN}{\mathcal{N}}
\newcommand{\cO}{\mathcal{O}}
\newcommand{\cS}{\mathcal{S}}
\newcommand{\Ed}[2]{\mathbb{E}_{#1}\left[#2\right]}
\usepackage{mathtools}
\DeclarePairedDelimiter{\sqne}{\|}{\|_2^2}
%\DeclarePairedDelimiter{\norme}{\|}{\|_2}
\DeclarePairedDelimiter{\normf}{\|}{\|_\mathrm{F}}
\DeclarePairedDelimiter{\normkfk}{\|}{\|_{\text{KF-}k}}
\DeclarePairedDelimiter{\normfstar}{\|}{\|_\mathrm{F*}}
\DeclarePairedDelimiter{\normftwo}{\|}{\|_\mathrm{F2}}
\DeclarePairedDelimiter{\sqns}{\|}{\|_{\mathrm{2}}^2}
\DeclarePairedDelimiter{\norms}{\|}{\|_{\mathrm{2}}}
\DeclarePairedDelimiter{\sqnn}{\|}{\|_{\mathrm{nuc}}^2}
\DeclarePairedDelimiter{\sqnf}{\|}{\|_{\mathrm{F}}^2}
\DeclarePairedDelimiter{\normn}{\|}{\|_{\mathrm{nuc}}}
\def\<#1,#2>{\langle #1,#2\rangle}
\DeclarePairedDelimiter{\dotprod}{\langle}{\rangle}

@@

Using this formula it is easy to compute updates for algorithms induced by various norms $\norm{\cdot}$. For example, when the norm $\norm{\cdot}$ is the Frobenius norm $\normf{\cdot}$, \cref{eq:our_update} turns into
\begin{equation}\label{eq:nsgd_update}
	\mX^{k+1} = \mX^{k} - \eta_k \frac{\mM^k}{\normf{\mM^k}}\,,
\end{equation} 
which recovers Normalized SGD (NSGD). When one has the spectral norm $\norms{\cdot}$, one gets
\begin{equation}\label{eq:muon_update}
	\mX^{k+1} = \mX^{k} - \eta_k U V^\top\,,
\end{equation} 
which is Muon without the $\sqrt{m/n}$ factor.
Here, $\mM^k = \mU \mSigma \mV^\top$ is the Singular Value Decomposition (SVD) of $\mM^k$ ($\mU = [u_1, u_2, \dots, u_r]$, $\mSigma = \diag(\sigma_1, \space \sigma_2, \dots, \sigma_r)$, and $\mV = [v_1, v_2, \dots, v_r]$). Muon can be recovered by taking the RMS-to-RMS operator norm $\sqrt{\frac{n}{m}}\norms{\cdot}$.

@@

\textbf{Ky Fan norms.}
There is another family of matrix norms, which might help us: Ky Fan norms. For $k \in \{1, \dots, \min(m, n)\}$, the Ky Fan $k$-norm $\normkfk{\cdot}$ is $\sum_{i=1}^{k}\sigma_i$, i.e. the sum of the $k$ largest singular values of the matrix. Special cases of the Ky Fan $k$-norm are the Ky Fan $1$-norm, which is the sprectral norm, and the Ky Fan $\min\{m, n\}$-norm, which is the nuclear norm.

@@

\begin{lemma}\label{lemma:dual_to_conv_comb}
	Let $\norm{\cdot}_{(1)},\dots,\norm{\cdot}_{(n)}$ be norms on a finite-dimensional Euclidean space, and let $\alpha_1,\dots,\alpha_n$ be non-negative reals. Define
	$$
		\norm{x} := \sum_{i = 1}^n \alpha_i \norm{x}_{(i)}.
	$$
	Then the dual unit ball of $\norm{\cdot}$ satisfies
	$$
		\cB_{\norm{\cdot}^\dagger}
		= \sum_{i = 1}^n \alpha_i \cB_{\norm{\cdot}_{(i)}^\dagger}
	$$
	where $\sum$ denotes the Minkowski sum and $\cB_{\norm{\cdot}_{(i)}^\dagger}$ is the unit ball of the dual norm $\norm{\cdot}_{(i)}^\dagger$.
\end{lemma}

@@

Applying it to optimization algorithms we obtain the following corollary.

@@

Combining Fanions-$k$ with another excellent LMO-algorithm, NSGD, we obtain the family of algorithms with updates defined by
\begin{equation}\label{eq:nsgd_muon_update}
		\mX^{t+1} = \mX^{t} - \gamma \left(\alpha \sum_{i = 1}^k u_i v_i^\top + (1-\alpha)\frac{\mM^k}{\normf{\mM^k}}\right)\,.
\end{equation}
By \cref{corollary:lin_comb_alg}, this is an LMO-algorithm corresponding to the unit ball of the norm $\norm{\cdot}_{\mathrm{F-KF-k}}$, given by
\begin{equation}\label{eq:fstar_norm}
	\norm{\cdot}_{\mathrm{F-KF-k}}^\dagger = \alpha \normkfk{\cdot}^\dagger + (1-\alpha)\normf{\cdot}^\dagger\,.
\end{equation}
We name the derived algorithms \emph{F-Fanions}. A bit of information and visualizations related to the $\normfstar{\cdot}$ norm is presented in the appendix (see \cref{eq:normfstardual}, \cref{fig:fstardual}).

Two edge members of this family, with $k = 1$ and $k = \min\{m, n\}$ correspondingly, are \emph{F-Neon} and \emph{F-Muon}.

\subsection{Add SignSGD: S-Muon and S-Neon}
Trivial formulas, but we need more experiments.

>>>>

%------------------------------------------------------------------------------------
% \epsilon is predefined; override to use \varepsilon
\renewcommand{\epsilon}{\varepsilon}
\newcommand{\Rmn}{\R^{m\times n}}
\newcommand{\cB}{\mathcal{B}}
\newcommand{\cD}{\mathcal{D}}
\newcommand{\cN}{\mathcal{N}}
\newcommand{\cO}{\mathcal{O}}
\newcommand{\cS}{\mathcal{S}}
\newcommand{\Ed}[2]{\mathbb{E}_{#1}\left[#2\right]}
\usepackage{mathtools}
\DeclarePairedDelimiter{\sqne}{\|}{\|_2^2}
%\DeclarePairedDelimiter{\norme}{\|}{\|_2}
\DeclarePairedDelimiter{\normf}{\|}{\|_\mathrm{F}}
\DeclarePairedDelimiter{\normkfk}{\|}{\|_{\text{KF-}k}}
\DeclarePairedDelimiter{\normfstar}{\|}{\|_\mathrm{F*}}
\DeclarePairedDelimiter{\normftwo}{\|}{\|_\mathrm{F2}}
\DeclarePairedDelimiter{\sqns}{\|}{\|_{\mathrm{2}}^2}
\DeclarePairedDelimiter{\norms}{\|}{\|_{\mathrm{2}}}
\DeclarePairedDelimiter{\sqnn}{\|}{\|_{\mathrm{nuc}}^2}
\DeclarePairedDelimiter{\sqnf}{\|}{\|_{\mathrm{F}}^2}
\DeclarePairedDelimiter{\normn}{\|}{\|_{\mathrm{nuc}}}
\DeclarePairedDelimiter{\normc}{\|}{\|_{\mathrm{C}}}
\def\<#1,#2>{\langle #1,#2\rangle}
\DeclarePairedDelimiter{\dotprod}{\langle}{\rangle}

@@

Using this formula it is easy to compute updates for algorithms induced by various norms $\norm{\cdot}$.

For example, when the norm $\norm{\cdot}$ is the Frobenius norm $\normf{\cdot}$, \cref{eq:our_update} turns into
\begin{equation}\label{eq:nsgd_update}
	\mX^{k+1} = \mX^{k} - \eta_k \frac{\mM^k}{\normf{\mM^k}}\,,
\end{equation} 
which recovers Normalized SGD (NSGD).

And when the norm is the spectral norm $\norms{\cdot}$, we get
\begin{equation}\label{eq:muon_update}
	\mX^{k+1} = \mX^{k} - \eta_k U V^\top\,,
\end{equation} 
which is Muon without the $\sqrt{m/n}$ factor.
Here, $\mM^k = \mU \mSigma \mV^\top$ is the Singular Value Decomposition (SVD) of $\mM^k$ ($\mU = [u_1, u_2, \dots, u_r]$, $\mSigma = \diag(\sigma_1, \space \sigma_2, \dots, \sigma_r)$, and $\mV = [v_1, v_2, \dots, v_r]$). Muon can be recovered by taking the RMS-to-RMS operator norm $\sqrt{\frac{n}{m}}\norms{\cdot}$.

When the norm is the Chebyshev norm $\normc{\cdot}$, we get
\begin{equation}\label{eq:signsgd_update}
	\mX^{k+1} = \mX^{k} - \eta_k \sign(\mM^k)\,,
\end{equation}
which recovers SignSGD~\citep{bernstein2018signsgd}. Here, $\sign(\mM^k)$ denotes the element-wise sign function. SignSGD is particularly notable for its communication efficiency in distributed training, as it compresses gradients to 1-bit per parameter.

@@

\textbf{Ky Fan norms.}
There is another family of matrix norms, which might help us: Ky Fan norms. For $k \in \{1, \dots, \min(m, n)\}$, the Ky Fan $k$-norm $\normkfk{\cdot}$ is $\sum_{i=1}^{k}\sigma_i$, i.e. the sum of the $k$ largest singular values of the matrix. Special cases of the Ky Fan $k$-norm are the Ky Fan $1$-norm, which is the spectral norm, and the Ky Fan $\min\{m, n\}$-norm, which is the nuclear norm.

@@

\begin{lemma}\label{lemma:dual_to_conv_comb}
	Let $\norm{\cdot}_{(1)},\dots,\norm{\cdot}_{(n)}$ be norms on a finite-dimensional Euclidean space, and let $\alpha_1,\dots,\alpha_n$ be non-negative reals. Define
	$$
		\norm{\cdot} := \sum_{i = 1}^n \alpha_i \norm{\cdot}_{(i)}.
	$$
	Then the dual unit ball of $\norm{\cdot}$ satisfies
	$$
		\cB_{\norm{\cdot}^\dagger}
		= \sum_{i = 1}^n \alpha_i \cB_{\norm{\cdot}_{(i)}^\dagger}
	$$
	where $\sum$ denotes the Minkowski sum and $\cB_{\norm{\cdot}_{(i)}^\dagger}$ is the unit ball of the dual norm $\norm{\cdot}_{(i)}^\dagger$.
\end{lemma}

@@

Applying it to optimization algorithms, we obtain the following corollary.

@@

Combining Fanions for various $k$ with another excellent LMO-algorithm, NSGD, we obtain the family of algorithms with updates defined by
\begin{equation}\label{eq:nsgd_muon_update}
		\mX^{t+1} = \mX^{t} - \gamma_k \left(\alpha \sum_{i = 1}^k u_i v_i^\top + (1-\alpha)\frac{\mM^k}{\normf{\mM^k}}\right)\,.
\end{equation}
Recall that Fanion-$k$ update is governed by the norm dual to the Ky Fan-$k$ norm, and NSGD corresponds to the Frobenius norm, which is dual to itself. Thus, by \cref{corollary:lin_comb_alg}, our new algorithm is an LMO-algorithm corresponding to the unit ball of the norm $\norm{\cdot}_{\mathrm{F-KF-k}}$:
\begin{equation}\label{eq:fkfk_norm}
	\norm{\cdot}_{\mathrm{F-KF-k}}^\dagger = \alpha \normkfk{\cdot} + (1-\alpha)\normf{\cdot}\,.
\end{equation}
We name the derived family of algorithm \emph{F-Fanions}.

Two edge members of this family, with $k = 1$ and $k = \min\{m, n\}$ correspondingly, are \emph{F-Neon} and \emph{F-Muon}, are of particular interest to us. A bit of information and visualizations related to the $\normfstar{\cdot} = \norm{\cdot}_{\mathrm{F-KF-1}}$ norm is presented in the appendix (see \cref{eq:normfstardual}, \cref{fig:fstardual}).

\subsection{Add SignSGD: S-Muon and S-Neon}

Another algorithm we consider is the SignSGD. Mixing its update with that of Fanion-$k$, we obtain the update of \emph{S-Fanion-$k$}:
\begin{equation}\label{eq:sign_muon_update}
		\mX^{t+1} = \mX^{t} - \gamma_k \left(\alpha \sum_{i = 1}^k u_i v_i^\top + (1-\alpha)\eta\sign(\mM^k)\right)\,,
\end{equation}
where $\eta$ is the special SignSGD learning rate coefficient.

The norm $\norm{\cdot}_{\mathrm{C-KF-k}}$ which produces this update in the LMO framework is again given by \cref{corollary:lin_comb_alg}:
\begin{equation}\label{eq:ckfk_norm}
	\norm{\cdot}_{\mathrm{C-KF-k}}^\dagger = \alpha \normkfk{\cdot} + \frac{1-\alpha}{\eta}\normc{\cdot}^\dagger\,.
\end{equation}
